---
title: "Project 1"
author: "Rachel David"
date: "2020-03-13"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

```{r}
#Data Sets:
library(tidyr)
head(population)
pop <-as.data.frame(population)

library(VGAM)
head(olym12)
olympics <- as.data.frame(olym12)
```

The two datasets I have chosen are 'population', which contains the variables: country, year, and population, and 'olym12', which contains the variables: ranking, country, gold, silver, bronze, and totalmedal. The population dataset was within the tidyr package and the olym12 dataset was in the VGAM package so I obtained this data through calling these libraries in R. I chose to look at these datasets as the Olympics has always been something I have loved watching, and I am interested to determine any associations between these datasets. I hypothesize that countries with larger populations will have a larger total medal count and that countries with larger overall total medal count will have more gold medals. 


```{R}
#Tidying: 
pop <- pop %>% pivot_wider(names_from ="year", values_from="population")

pop1 <- pop[-c(2:18, 20)]
```

To tidy my pop data, I used pivot_wider so each country had one row and the years were moved to columns. I did this as the other data set has only one row per country, and this will allow for an easier join and so that each observation could have its own row and each variable had its own column. I then edited the data further so that all year columns (18 in total) were dropped in the population dataset except for the year 2012, as this is the one year that the olympic data was collected from. 

```{R}
#Joining
library(dplyr)
as.data.frame(apply(olympics,2,function(x)gsub('\\s+', '',x)))
pop1 <- as.data.frame(apply(pop1,2,function(x)gsub('\\s+', '',x)))
names(pop1)[names(pop1) == "2012"] <- "popul"
pop1$popul <- as.numeric(as.character(pop1$popul))
pop1 %>% mutate_if(is.factor, as.character) -> pop1

test <- pop1
pop1$country <- gsub("China,HongKongSAR", "HongKong", pop1$country)
pop1$country <- gsub("UnitedStatesofAmerica", "UnitedStates", pop1$country)
pop1$country <- gsub("UnitedKingdomofGreatBritainandNorthernIreland", "GreatBritain", pop1$country)
pop1$country <- gsub("RussianFederation", "Russia", pop1$country)
pop1$country <- gsub("RepublicofKorea", "SouthKorea", pop1$country)
pop1$country <- gsub("DemocraticPeople'sSouthKorea", "NorthKorea", pop1$country)
pop1$country <- gsub("Iran\\(IslamicRepublicof\\)", "Iran", pop1$country)
pop1$country <- gsub("Venezuela\\(BolivarianRepublicof\\)", "Venezuela", pop1$country)
pop1$country <- gsub("VietNam", "Vietnam", pop1$country)
outlier <- olympics %>% anti_join(pop1,by=c("country"="country"))%>% distinct(country)
outlier
outlier2 <- pop1 %>% anti_join(olympics,by=c("country"="country"))%>% distinct(country)
glimpse(outlier2)
pop1$country <- gsub("RepublicofMoldova", "Moldova", pop1$country)

total <- inner_join(pop1, olympics, by=c("country"="country"))
head(total)


```
I did an inner join to join my two datasets, as the olympics dataset had much fewer countries than the 
populations dataset as many countries did not participate in the 2012 Olympics. This introduced many NAs, 
which would result in issues when doing data wrangling, so I did an inner join so only the countries present 
in both datasets would be included. This resulted in 1 row being dropped from the olympics dataset and 135 
rows being dropped from the populations dataset. This was a large number of cases dropped from the populations 
dataset as many countries did not participate in the Olympics, and this should be noted as much data was lost. 
    I also had to do some manipulation to the names as the countries were not named exactly the same (one 
dataset had spaces while others didn't and some used abbreviations that the other did not), so I used gsub() 
to adjust these names so these elements could be included.
      

```{r}
##Wrangling
library(dplyr)
library(ggplot2)

total %>% select(country, popul, totalmedal)  %>% filter(totalmedal>10 & popul>10000000)

total %>% select(country,gold) %>% arrange(desc(gold))

total <- total %>% mutate(poppctile=ntile(popul,100))
glimpse(total)

total <-total %>% mutate (size = case_when(popul<10000000 ~ "small", 
popul<=50000000 & popul>=10000000 ~ "medium", popul>50000000 ~ "big"))

total %>% group_by(size) %>% summarize(mean_medal=mean(totalmedal))

##Summary Statistics 
total %>% group_by(size) %>%summarize(max(totalmedal))
total %>% group_by(size) %>%summarize(max(gold))
total %>% group_by(size) %>%summarize(mean(ranking))
total %>% group_by(size) %>%summarize(mean(popul))
total %>% group_by(size) %>%summarize(mean(gold))
total %>% summarize(sd(popul))
total %>% group_by(size) %>%summarize(median(popul))
total %>% group_by(size) %>%summarize(max(popul))
total %>% group_by(size) %>%summarize(min(totalmedal))
total %>% group_by(size) %>%summarize(sd(totalmedal))
total %>% summarize_all(n_distinct)

#

stats <- total %>% summarize_if(is.numeric, list(min=min,max=max, mean=mean, 
sd=sd,var=var, median=median))

stats%>%pivot_longer(contains("_"))%>%separate(name,into=c("variable","stat")) %>%
  pivot_wider(names_from = "variable", values_from = "value") %>% 
  arrange(stat) %>% tidyr::unnest()

groupstats <- total %>% group_by(size) %>% summarize_if(is.numeric, list(min=min,max=max, 
mean=mean, sd=sd,var=var, median=median))
groupstats%>%pivot_longer(contains("_"))%>%
  separate(name,into=c("variable","stat")) %>% 
  pivot_wider(names_from = "stat", values_from = "value") %>% 
  arrange(variable) %>% tidyr::unnest()


#corelation matrix 
corr <- total %>% select_if(is.numeric)
cor(corr)

```
   With the filter() function, I filtered the data to create a list of countries that received more than 10 medals total at the Olympics and had a "big" population; using the select() function, I was able to display only the country, totalmedal, and popul columns to make the results more clear.  Using the arrange() function, we see that the US was the country with the most number of gold medals. Using a dplyr vector function of ntile() with mutate(), we were able to create a new variable that showed what percentile (out of 100) that the country was in terms of its population. I then used mutate() to create a new categorical variable "size" to distinguish if the country was big, medium, or small in terms of its population size. I then did group_by() on this size and summarize() to determine the mean number of total medals received by each grouping of country size. We saw that big countries had the most medals followed by medium and finally small countries. Using n_distinct(), we can see the total number of distint elements in each variable - the most notable observation is that multiple countries have the same ranking as they received the same number of total medals. 
   
Using summary statistics, we see that after grouping by size, big countries have the most total medals followed by medium and then small; we see this same trend when grouping by size and looking at number of gold medals, overall ranking, and overall population size (descending from big to small). We then determined the standard deviation of the population of the countries; we grouped by size of country and found the median population value; we determined the largest population size in each size grouping of country; using summarize() and min(), we found the smallest number of total medals in each size grouping; we used summarize() and sd() to find the standard deviation in total medals when grouping by size. I created tables of both each numeric variable and each numeric variable grouped by country size that showed the min, max, median, mean, sd, and var. From these tables we see that in general, the number of medals increased as country size increased, but the minimum for all sizes was very comparable. I then created a correlation matrix of the numeric variables, and we can see stronger correlations between each type of medal and some correlation between overall population and medals received.   

```{r}
#Visualizing
library(ggplot2)
library(dplyr)
library(tidyverse)

cleancorr<-cor(corr)%>%as.data.frame%>%
  rownames_to_column%>%
  pivot_longer(-1,names_to="name",values_to="correlation")
head(cleancorr)

cleancorr %>% ggplot(aes(rowname,name,fill=correlation))+
  geom_tile()+
  scale_fill_gradient2(low="red",mid="white",high="blue")+ 
  geom_text(aes(label=round(correlation,2)),color = "black", size = 4)+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+ 
  coord_fixed()

#scatterplot 
ggplot(total, aes(x=popul, y=totalmedal, color=size))+
  geom_point() + 
  scale_y_continuous(name="Total Medals", breaks = seq(from = 0, to = 100, by = 20))+
  scale_color_brewer(palette="Set1")+
  labs( title="Total Medals per Country Size", x="Country Population")+ 
  theme(axis.text.x = element_text(angle=45, hjust=1), 
        legend.position="bottom")+scale_x_log10()



##bar graph 
total %>% group_by(size) %>% ggplot(aes(x=size,y=gold, fill=size)) +
  geom_bar(aes(x=size, y=gold,fill=size), stat="summary", fun.y="mean") + 
scale_fill_manual("legend", values = c("big" = "lightblue1", "medium" = "plum", "small" = "lightpink1"))+
  labs( title="Average Gold Medals per Country Size")+scale_y_continuous(name="Gold Medals")+
  scale_x_discrete(name ="Country Population Size")+ 
  theme( legend.position = c(.95, .95),legend.justification = c("right", "top"), 
         legend.box.just = "right",legend.margin = margin(6, 6, 6, 6)) +
  geom_errorbar(stat="summary",fun.data="mean_se", width=0.2)
 


```
From the correlation heatmap, we see a strong correlation between totalmedal and bronze, silver,
and gold. We also see an increasing correlation in population and medals as you increase from bronze to silver to gold. It was interesting to observe that there was not a very high correlation between population and medals in general as compared with correlation between total medals and each individual medal. We can see negative correlations between ranking and gold, silver, bronze, and totalmedals as the better a country is and more medals they receive, the higher (or smaller number) their rank becomes.

From the scatterplot, we see a general positive correlation that as population increases, the number of total medals also increases. However, there are countries of all groupings (big, medium, and small) that had very low total medals, so this is not a direct correlation. We see within each size grouping, there is also a general correlation bewtween an increase in population and an increase in total medals. After 40 medals total, we see exclusively big countries while, no small country has above 20 medals, and medium countries has a range of medals from 0-around 40. 

From the bar plot, we can clearly see that as the size grouping of the country increases, the average number of gold medals received also increases. This shows a direct positive correlation between country population size and the average number of gold medals received. We see a notable increase in gold meadls from small to medium and a much more drastic jump in gold meadls when moving from medium to big countries. We also can see much larger standard error for the big countries when compared to medium and small size countries. 
```{r}
#Dimentionality Reduction


library(cluster)
pam2p <- total %>% select(-country, -size) %>% scale %>% pam(2) ##scale then pam with 2 clusters  
pam2p

sil_width<-vector() ##check clusters with sil_width and graph 
for(i in 2:10){   
  pamp_fit <- total%>%select(-country, -size)%>%pam(i)  
  sil_width[i] <- pamp_fit$silinfo$avg.width  
} 

ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)

finalp<-total%>%mutate(cluster=as.factor(pam2p$clustering)) ##add cluster data 

confmatp <- finalp%>%group_by(size)%>%count(cluster)%>%
arrange(desc(n))%>% 
  pivot_wider(names_from="cluster",values_from="n",values_fill = list('n'=0)) 
##group to check
confmatp

ggplot(finalp, aes(x=popul,y=totalmedal, color=cluster))+geom_point() ##plot on 2 

library(GGally) 
ggpairs(finalp, columns=2:8, aes(color=cluster)) ##plot all pairs 


plot(pam2p,which=2) ##plot avg. sil width 

```
      I first scaled my numeric variables in the dataset. I then graphed the sil_width of all numeric variables and the graph suggested a total of 2 clusters. I then ran the cluster analysis with pam() using two clusters as determined on all numeric variables of my dataset. I then tested how accurate the groupings were by grouping by cluser and comparing it to the three size groupings, the clusters were exactly right for small size, not as accurate for medium and big size. The second cluster had much fewer in the cluster than the first. I then graphed on two variables (popul and totalmedal) in clusters based on all 7 numeric variables. 
      I then visualized all pairwise combinations of all the variables and found strong correlations that as ranking decreased (which means the country was doing better), the number of total medals was increasing. The clusters mapped well and I  saw strong correlations between the number of gold and silver medals as well as gold, silver, and bronze and total medals. The plots that showed the clusters most distinctly included the plots with ranking and gold, silver, bronze, and total medals. This makes sense as these variables would be correlated. I then interpreted the average silhouette width and it was 0.54, which shows that a reasonable structure has been found. 

```{R}